Legacy Data Confounds Genomics Studies

### 2019-04-04 legacy.tex
complete.

The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project (1kGP)\citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses. 
Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.

The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved.  
The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012} \sgcomment{also cite phase 3}, leading to heterogeneous sample preparations and data quality.

@@

To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure \citep{song2015testing}. \sgcomment{We also considered two alternative approaches to account for confounders, namely using the leading five  principal component analysis, and using population labels as covariates.   }

This method identifies a total of 24,390 variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}. We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
Lists of suspicious variants and individuals with low $Q$ are provided as supplementary material \luke{link to file here}.

@@

We did a similar analysis using the variants identified in the \textit{5PC} \sgcomment{I don't think that this has been described? I also don't understand why the 5pc approach has so many more variants than the logistic factor analysis above} model present in the 83 resequenced individuals.
Of the 24,390 suspicious variants identified globally, only 12,086 of them are found in the 83 individuals in the original 1kGP data (See Figure \ref{90HanSFS_full}).
From this subset, only 1,637 of these variants are present in the resequenced data, the rest are missing entirely.
Figure \ref{90HanSFS_full} shows a relatively weak correlation to allele frequency for the variants present in the resequenced data.
A number of variants shared across cohorts suggests that roughly 10\% of the variants identified are present in the cell lines real but exhibit genotyping bias.

@@

\subsubsection{Imputation}

@@

\subsubsection{Microarrays}
\sgcomment{It seems like this paragraph is out of place and would fit better in the validation section, above? }
We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip. This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position. Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.

\subsubsection{GWAS}
\sgcomment{"these dubious", below, refers to all dubious positions rather than only the ones on the chips, right? Even more reason to move the previous paragraph. }

@@

All papers used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
They also removed rare alleles and alleles with high degrees of missingness.
Despite using state-of-the-art quality controls, these erroneous \sgcomment{we don't know that all or even most are erroneous} variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for association with biological traits.

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations. To distinguish between weak but significant biases and dubious SNPS, we also considered the effect size and distinguished between variants where the allele frequency difference between low- and high-$Q$ individuals is larger than a factor of two. \sgcomment{you know, we could simply compare the relative effect size to predicted frequency $\pi$}

\begin{table}[h]
\begin{tabular}{l l l r r r}
  {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$ -$\log_{10} p$\\(adjusted)}      \\ \hline
28654678 & PLoS One & rs201761909 & 5.7 & 78.11\\
	& 	&  	rs201130852 & 5.05 & 72.28\\
	& 	&  	rs201255786 & 5.7 & 68.97\\
	& 	&  	rs200655768 & 6.52 & 66.67\\
	& 	&  	rs184202621 & 5.52 & 60.45\\
	& 	& 	rs80274284 & 6 & 56.15\\
	& 	& 	rs200699422 & 5.3 & 7.43\\
23527680 & \multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics} & rs6057648 & 5.4 & 20.5\\
28928442 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & rs201471471 & 6.52 & 7.87\\
26053186 & \multicolumn{1}{p{3cm}}{\raggedright PLoS One} & rs60136336 & 5.7 & 2.25\\ \hline
\textcolor{gray}{28270201} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Genome Medicine}} & \textcolor{gray}{rs453755} & \textcolor{gray}{7.52} & \textcolor{gray}{5.29}\\
\textcolor{gray}{23023329} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*15.3} & \textcolor{gray}{4.32}}\\
\textcolor{gray}{28334899} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*29.3} & \textcolor{gray}{4.32}\\
\textcolor{gray}{28240269} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Communications}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*72.7} & \textcolor{gray}{4.32}\\
\textcolor{gray}{27863252} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Cell}\\} & \textcolor{gray}{rs3794738} & \textcolor{gray}{*13.15} & \textcolor{gray}{3.73}\\
\textcolor{gray}{29534301} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Hepatology}\\} & \textcolor{gray}{rs9273062} & \textcolor{gray}{*9.7} & \textcolor{gray}{3.36}\\
\textcolor{gray}{21386085} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Diabetes}\\}  & \textcolor{gray}{rs301} & \textcolor{gray}{*10.52} & \textcolor{gray}{3.02}\\
\textcolor{gray}{26830138} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Molecular Psychiatry}} & \textcolor{gray}{rs77894924} & \textcolor{gray}{6.7} & \textcolor{gray}{2.77}\\
\textcolor{gray}{29617998} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs4963156} & \textcolor{gray}{*22.4} & \textcolor{gray}{2.52}\\
\textcolor{gray}{28698626} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Scientific Reports}} & \textcolor{gray}{rs11015915} & \textcolor{gray}{5.05} & \textcolor{gray}{2.45}\\
\textcolor{gray}{26974007} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs3124998} & \textcolor{gray}{*8.05} & \textcolor{gray}{2.33}\\
\textcolor{gray}{26634245} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{BMC Genetics}} & \textcolor{gray}{rs451000} & \textcolor{gray}{6} & \textcolor{gray}{2.28}\\
 	& 	& 	\textcolor{gray}{rs443874} & \textcolor{gray}{5.3} & \textcolor{gray}{2.26}\\
	& 	&	\textcolor{gray}{rs400942} & \textcolor{gray}{6} & \textcolor{gray}{2.2}\\
\textcolor{gray}{25918132} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Toxicological Sciences}} & \textcolor{gray}{rs76780579} & \textcolor{gray}{6} & \textcolor{gray}{2.09}\\

@@

\subsection{Technical Artifact}
The variants identified in this study are likely to be technical artifacts from legacy technologies.
Different sequencing technologies will have different error profiles. 
A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011} with, for instance, different patterns of quality decrease along reads. %(known as B-tails). 
Differences in read quality and error profiles in turn require different calling pipelines.

@@

%For instance, our test was limited by the genome wide resolution of the quality per sample.
%Data quality at a SNP-wise resolution would be a better metric to use, but unfortunately was not available.

%If the Illumina Genome Analyzer sequencing platforms are no longer being used, then it it likely that these sequencer specific artifacts are no longer 
However, because the 1kGP data is widely used as a reference database, these variants are still being imputed onto new genotype data and can then impact association studies for a variety of phenotypes. The significant association of a variant with a quality metric is not in itself an indication that the variant is spurious. It may simply be subject to a weak but significant calling bias --- not an extraordinary surprise in a low-coverage dataset with a complex, joint sample variant calling pipeline. It may nevertheless be prudent to discard such variants from analyses that may be sensitive to calling biases, such as mutation profiles and allele frequency estimates. The imputation of such variants can also lead to milder problems such as the addition of noise in polygenic risk scores, biases in fine mapping analyses.

%Even though we expect that present-day technologies will be less sensitive to such artefacts, it is worth pointing out that cancer mutational signature analysis can also be sensitive to technological artefacts that are difficult to detect using classical quality metrics.

%For conventional GWAS, we do not see evidence that these contaminated hits impact the conclusions made in each of these publications.
%However, considering many polygenic risk score estimates use the association scores from multiple sources, the inclusion of these variants could possibly impact downstream analyses.

\subsection{Recommendations}
While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet quality standards expected from high-quality sequencing data.
In order to avoid spurious results, the most conservative approach would be to remove or resequence all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$. 
We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signals (Supplementary Figure \ref{MutSpect}).
Another approach, for researchers relying on imputation servers where reference panel individuals cannot be filtered out, is to remove sites that reach a significance below $ p > -\log_{10}(\alpha)$ in the reverse GWAS. (The supplementary material contains a list of variants statistically associated to $Q$).

@@

\subsection{Data Availability}

@@

We run a total of $S$ regressions, where $S$ is the total number of genomic loci. Given the large number of tests, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. 
See supplementary file {here}. \luke{make this file}

@@

%\subsection{Frequencies of variants in other datasets}
%Using the allele frequency files for the Japanese HapMap samples, we extracted the positions we identified as being significantly associated to the average quality of mapped bases. We then compared the frequencies of these 70 variants in the 1kGP Japanese samples to those of the HapMap Japanese samples. Figure \ref{HapMap_GnomAD} compares the allele frequencies for suspicious sites that are present in HapMap.

%Similarly, using the allele frequencies from GnomAD, we were able to extract 186 sites that are statistically significantly associated to the average quality of mapped bases. However, in this case we compared global allele frequencies in both the 1kGP and GnomAD (Figure \ref{HapMap_GnomAD}).

\section{Code Availability}
https://github.com/LukeAndersonTrocme/QualityPaper

>>>>

The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project (1kGP)\citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012,phase3}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses. 
Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.

The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved.  
The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012,phase3}, leading to heterogeneous sample preparations and data quality.

@@

To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}. 
We also considered two alternative approaches to account for confounders, namely using the leading five  principal component analysis, and using population labels as covariates (Supplementary Figure \ref{CompareModel}).

This method identifies a total of 24,390 variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}. 
We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
Lists of suspicious variants and individuals with low $Q$ are provided as supplementary material \ref{rsIDList}.

@@

We did a similar analysis using the variants identified in the GCAT model present in the 83 resequenced individuals.
Of the 24,390 suspicious variants identified globally, only 12,086 of them are found in the 83 individuals in the original 1kGP data (See Figure \ref{90HanSFS_full}).
From this subset, only 1,637 of these variants are present in the resequenced data, the rest are missing entirely.
Figure \ref{90HanSFS_full} shows a relatively weak correlation to allele frequency for the variants present in the resequenced data.
A number of variants shared across cohorts suggests that roughly 10\% of the variants identified are present in the cell lines real but exhibit genotyping bias.

We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip. 
This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position. 
Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.

@@

@@

@@

All papers used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
They also removed rare alleles and alleles with high degrees of missingness.
Despite using state-of-the-art quality controls, these variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for association with biological traits.

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations. To distinguish between weak but significant biases and dubious SNPS, we also considered the effect size and distinguished between variants where the allele frequency difference between low- and high-$Q$ individuals is larger than a factor of two.

\begin{table}[h]
\begin{tabular}{l l l r r r}
  {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$ -$\log_{10} p$\\(adjusted)}      \\ \hline
28654678 & PLoS One & rs201761909 & 5.7 & 78.11\\
	& 	&  	rs201130852 & 5.05 & 72.28\\
	& 	&  	rs201255786 & 5.7 & 68.97\\
	& 	&  	rs200655768 & 6.52 & 66.67\\
	& 	&  	rs184202621 & 5.52 & 60.45\\
	& 	& 	rs80274284 & 6 & 56.15\\
	& 	& 	rs200699422 & 5.3 & 7.43\\
23527680 & \multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics} & rs6057648 & 5.4 & 20.5\\
28928442 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & rs201471471 & 6.52 & 7.87\\
26053186 & \multicolumn{1}{p{3cm}}{\raggedright PLoS One} & rs60136336 & 5.7 & 2.25\\ \hline
\textcolor{gray}{28270201} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Genome Medicine}} & \textcolor{gray}{rs453755} & \textcolor{gray}{7.52} & \textcolor{gray}{5.29}\\
\textcolor{gray}{23023329} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*15.3} & \textcolor{gray}{4.32}\\
\textcolor{gray}{28334899} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*29.3} & \textcolor{gray}{4.32}\\
\textcolor{gray}{28240269} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Communications}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*72.7} & \textcolor{gray}{4.32}\\
\textcolor{gray}{27863252} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Cell}\\} & \textcolor{gray}{rs3794738} & \textcolor{gray}{*13.15} & \textcolor{gray}{3.73}\\
\textcolor{gray}{29534301} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Hepatology}\\} & \textcolor{gray}{rs9273062} & \textcolor{gray}{*9.7} & \textcolor{gray}{3.36}\\
\textcolor{gray}{21386085} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Diabetes}\\}  & \textcolor{gray}{rs301} & \textcolor{gray}{*10.52} & \textcolor{gray}{3.02}\\
\textcolor{gray}{26830138} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Molecular Psychiatry}} & \textcolor{gray}{rs77894924} & \textcolor{gray}{6.7} & \textcolor{gray}{2.77}\\
\textcolor{gray}{29617998} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs4963156} & \textcolor{gray}{*22.4} & \textcolor{gray}{2.52}\\
\textcolor{gray}{28698626} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Scientific Reports}} & \textcolor{gray}{rs11015915} & \textcolor{gray}{5.05} & \textcolor{gray}{2.45}\\
\textcolor{gray}{26974007} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs3124998} & \textcolor{gray}{*8.05} & \textcolor{gray}{2.33}\\
\textcolor{gray}{26634245} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{BMC Genetics}} & \textcolor{gray}{rs451000} & \textcolor{gray}{6} & \textcolor{gray}{2.28}\\
 	& 	& 	\textcolor{gray}{rs443874} & \textcolor{gray}{5.3} & \textcolor{gray}{2.26}\\
	& 	&	\textcolor{gray}{rs400942} & \textcolor{gray}{6} & \textcolor{gray}{2.2}\\
\textcolor{gray}{25918132} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Toxicological Sciences}} & \textcolor{gray}{rs76780579} & \textcolor{gray}{6} & \textcolor{gray}{2.09}\\

@@

The variants identified in this study are likely to be technical artifacts from legacy technologies.
Different sequencing technologies will have different error profiles. 
A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011} with, for instance, different patterns of quality decrease along reads. 
Differences in read quality and error profiles in turn require different calling pipelines.

@@

However, because the 1kGP data is widely used as a reference database, these variants are still being imputed onto new genotype data and can then impact association studies for a variety of phenotypes. The significant association of a variant with a quality metric is not in itself an indication that the variant is spurious. It may simply be subject to a weak but significant calling bias --- not an extraordinary surprise in a low-coverage dataset with a complex, joint sample variant calling pipeline. It may nevertheless be prudent to discard such variants from analyses that may be sensitive to calling biases, such as mutation profiles and allele frequency estimates. The imputation of such variants can also lead to milder problems such as the addition of noise in polygenic risk scores, biases in fine mapping analyses.

While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet quality standards expected from high-quality sequencing data.
In order to avoid spurious results, the most conservative approach would be to remove or resequence all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$. 
We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signals (Supplementary Figure \ref{MutSpect}).
Another approach, for researchers relying on imputation servers where reference panel individuals cannot be filtered out, is to remove sites that reach a significance below $ p > -\log_{10}(\alpha)$ in the reverse GWAS. (The supplementary material contains a list of variants statistically associated to $Q$).

@@

\subsection{Code and data availability}
For more details regarding data processing or figure plotting, our pipelines and R Notebooks are publicly available \href{https://github.com/LukeAndersonTrocme/QualityPaper}{here}.

@@

We run a total of $S$ regressions, where $S$ is the total number of genomic loci. Given the large number of tests, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. 
See supplementary file  \ref{rsIDList}.

@@
